---
title: "Math E-23C Term Project"
subtitle: "An analysis of the Lakers 2019 Championship season"
author: "Maria Cristina An, Daniel Lebedinsky, Julio Solis Arce"
date: "`r format(Sys.time(), '%d %B, %Y')`"
abstract: "The current documents offer a comprehensive analysis of the Lakers 2019 season, in which they successfully defended the NBA championship"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Packages used
library(dplyr)
```

\newpage

# Dataset description

## Codebook

### Game level variables

- Rk -- Rank

- G -- Season Game

- W_E -- opponent is from East or West

- Opp -- Opponent

- W_L -- Lakers team won or lost

- Tm -- Lakers team Points

- Opp_pts -- Opponent Points

### Lakers relevant variables

- FG_LA -- Field Goals

- FGA_LA -- Field Goal Attempts

- FGpc_LA -- Field Goal Percentage

- X3P_LA -- 3-Point Field Goals

- X3PA_LA -- 3-Point Field Goal Attempts

- X3Ppc_LA -- 3-Point Field Goal Percentage

- FT_LA -- Free Throws

- FTA_LA -- Free Throw Attempts

- FTpc_LA -- Free Throw Percentage

- ORB_LA -- Offensive Rebounds

- TRB_LA -- Total Rebounds

- AST_LA -- Assists

- STL_LA -- Steals

- BLK_LA -- Blocks

- TOV_LA -- Turnovers

- PF_LA -- Personal Fouls

### Opponent relevant variables

- FG -- Opponent Field Goals

- FGA -- Opponent Field Goal Attempts

- FGpc -- Opponent Field Goal Percentage

- X3P -- Opponent 3-Point Field Goals

- X3PA -- Opponent 3-Point Field Goal Attempts

- X3Ppc -- Opponent 3-Point Field Goal Percentage

- FT -- Opponent Free Throws

- FTA -- Opponent Free Throw Attempts

- FTpc -- Opponent Free Throw Percentage

- ORB -- Opponent Offensive Rebounds

- TRB -- Opponent Total Rebounds

- AST -- Opponent Assists

- STL -- Opponent Steals

- BLK -- Opponent Blocks

- TOV -- Opponent Turnovers

- PF -- Opponent Personal Fouls

## Importing and cleaning data
Reference: https://www.basketball-reference.com/teams/LAL/2020/gamelog/
```{r, message = FALSE, warning = FALSE}

#Import dataset
Lakers <- read.csv("Lakers\ 2019-20\ Game\ log.csv"); head(Lakers, n = c(6, 8)) %>% knitr::kable()

# Convert to proper date format
Lakers$Dates <- as.Date(Lakers$Date)
attach(Lakers)
```

## Summary of variables

```{r}
summary(Lakers) %>%
  t() %>%
  as.data.frame() %>%
  select(-Var2) %>%
  rename(Variable=Var1, Description=Freq) %>%
  filter(!is.na(Description)) %>%
  arrange(Variable) %>%
  knitr::kable()
```

## Number of columns and rows in dataset
```{r}
length(Lakers)
nrow(Lakers)
# Creating categorical columns
WonLost <- W_L == "W"; head(WonLost) 
sum(WonLost) # total wins
```

\newpage

# Graphical analysis

## Barplot of wins vs losses for our team
 This plot shows the number of losses compared to the number wins by our team. 
 The y-axis gives the count of wins and losses. The x-label answers the question
 "did the team win?", FALSE means losing and TRUE means winning.
```{r}
library(ggplot2)
Lakers %>% ggplot(aes(WonLost))+geom_bar(fill="orange")+
  xlab("Won Game?") +
  ggtitle("Total Lakers Wins and Losses")
```
 

## Barplot of 3 point shots from our team

This plot shows the number of times the team scores a 3-point shot in a game.
How often a 3-pointer occurs ranges from the minimum of 2 to a maximum of 19
in a single game.
```{r}
Lakers %>% ggplot(aes(X3P_LA))+geom_bar(fill="yellow")+
  xlab("3-pointer count") +
  ylab("Frequency")+
  ggtitle("Lakers' Number of 3-Pointers per Game")
```


## Barplot of Opponent blocks 

This plot shows how often an opponent successfully blocks our team in a game.
Most often (the median), an opponent is able to block 3 times and, of course, 
depending on the opponent's abilities, blocks can range from 0 to 9 in a game.
```{r}
Lakers %>% ggplot(aes(BLK))+geom_bar(fill="green")+
  scale_x_continuous(breaks=c(1:10))+
  xlab("Block count") +
  ylab("Frequency")+
  ggtitle("Lakers' Number of Opponent Blocks per Game") 
```


## Histogram of 3 point shots

This is a histogram version of the barplot which shows the number of times the team scores a 3-point shot in a game.
The data is the same, and so is the result, a 3-pointer event ranges from the minimum of 2 to a maximum of 19.
```{r}
Lakers %>%
  ggplot(aes(X3P_LA)) +
  geom_histogram(binwidth = 1, fill="cyan", col = "black")+
  scale_x_continuous(breaks=c(0:20))+
  ylab("Frequency")+
  xlab("3 point shots")+
  ggtitle("Histogram of Lakers' 3 Pointers") 
```


## Histogram of Opponent blocks
 
This is a histogram version that shows how often an opponent successfully blocks our team in a game.
Same results, an opponent is able to block 3 times and ranges from 0 to 9 in a game.
```{r}
Lakers %>%
  ggplot(aes(BLK)) +
  geom_histogram(binwidth = 1, fill="magenta", col = "black")+
  scale_x_continuous(breaks=c(0:12))+
  ylab("Frequency")+
  xlab("Number of Blocks")+
  ggtitle("Histogram of Lakers' Opponent Blocks") 
```


## Histogram of our team's number of Free Throws

This histogram shows the probability of the number of free throws done by our team in a game.
The range is from a minimum of 7 to a maximum of 33 with a maximum density at 14 and 17.
```{r}
hist(FT_LA, probability=TRUE, main="Histogram of Lakers' Number of Free Throws per Game", xlab="Number of Free Throws", col = "pink", breaks=(0:35)) 
# free throws from the team per game
```

\newpage

# Probability analysis

## Probability density graph overlay on the our team's total number of rebounds in a game

Here we have our team's probability of the total number of rebounds in a game.
Ranging from a minimum of 25 to a maximum of 62, the graph seems like it came from a 
normal distribution. Overlaying a normal distribution shows like it does.
```{r}
mu<-mean(TRB_LA); stdv<- sd(TRB_LA)
hist(TRB_LA, probability=TRUE, main = "Histogram of Lakers' Total Number of Rebounds per Game", ylim=c(0,0.08), xlab="Number of Rebounds", col = "pink", breaks=(20:65)) 
# total rebounds per game  
curve(dnorm(x, mu, stdv), col="purple", add=TRUE, lwd=2) 
#Appears to be a normal distribution. We will compare it with Deciles and a Q-Q plot
Normal10<- qnorm(seq(0.01, 0.99, by=0.1), mean=mu, sd=stdv);
Data10<- quantile(TRB_LA, seq(0.01, 0.99, by=0.1), type=2);
plot(Normal10, Data10)
f<-function(x) x
curve(f, col="red", add=TRUE)
#Our data does not differ much from the Normal Distribution, except in the first decile.
```

## Probability density graph on the our team's scores in a game

We go back to our histogram of our team's score and its probability in a game.
It also seems as if it fits a normal distribution like it shows with a normal distribution overlay.
We can provide more proof later with a chi-square goodness of fit test.
```{r}
mu<-mean(Tm); stdv<-sd(Tm)
hist(Tm,breaks=seq(from=80, to = 150, by =5), probability=TRUE, main = "Histogram of Lakers' Scores per Game", xlab="Scores", col = "light blue") # team's score frequencies
curve(dnorm(x, mu, stdv), col="dark blue", add=TRUE, lwd=2)

#Binning and Chisq test to compare this histogram to a Normal Distribution
dec <- qnorm(seq(0.0, 1, by = 0.1), mu, stdv); dec   #10 bins
Exp <- rep(length(Tm)/10,10); Exp     #expected scores per bin
#Now we can count how many scores are in each bin
binscores <- numeric(10)
for (i in 1:10)  
  binscores[i] <- sum((Tm >= dec[i]) & (Tm <= dec[i+1]) ); binscores
#Test for uniformity using chi square.
Chi2 <- sum((binscores - Exp)^2/Exp); Chi2
#There were 10 bins. We estimated 2 parameters (mu and stdv), which costs two degrees of freedom
#Also we "made the totals match", costing another 1. So there are 10-2-1=7 df.
curve(dchisq(x, df = 7), from = 0, to = 120 )
abline(v=Chi2, col = "red")
#The probability of this chi-square value is relatively large
#The normal distribution was a good model
```

## Contingency tables

This table answers the question "how many times the opponents scored greater than 100 points in a game with our team".
This does not say that the opponent won or not. But in the times that they did not, it only shows that our team scored more and that such a high scoring game would have been a truly awesome event to watch.
```{r}
Opp100 <- Opp_pts > 100; Opp100
table(Opp100)
```

This table shows the opposite of the table above, that despite being the all-around champion during that year, our team scored less than 100 in some games and this table shows that.
```{r}
Less100 <- Tm < 100; Less100
table(Less100)
```

\newpage

# Inferential and statistical analysis

## Permutation test
Our permutation test will be comparing our team's scores in games played vs the opponents from the East and the opponents from the West

H_o: The mean score of the opponents coming from the Eastern Conference is equal
to the mean score of the opponents coming from the Western Conference
H_a: The mean score of the opponents coming from the Eastern Conference is not equal
to the mean score of the opponents coming from the Western Conference

#Data for Eastern and Western Conference teams
How many opponents were from each, and what were their mean scores? 
```{r}
Eastern <- W_E == "E"; head(Eastern) 
East <- sum(Eastern); East # count of Eastern opponents = 31
ScorevsEast <- mean(Tm*Eastern); ScorevsEast # mean team score vs East = 37.03261

Western <- W_E == "W"; head(Western) 
West <- sum(Western); West # Western opponents = 61
ScorevsWest <- mean(Tm*Western); ScorevsWest # mean team score vs West = 76.26087
```
It shows that there are more opponents coming from the West and that their mean score
obviously would have a higher range.

## Total number of games played and differences of mean scores 
This will be our Observed value.
```{r}
EW <- sum(Eastern) + sum(Western); EW # total number of games = 92
Score_diff <- ScorevsEast - ScorevsWest; Score_diff # -39.22826
Observed <- Score_diff; Observed
# Let's see if this score difference is significant
# We repeat 10^6 times
N <- 10^6
Score_diffs <- numeric(N)
for (i in 1:N){
  # Permute West indices
  E <- sample(EW, East, replace = FALSE)
  
  # Get the difference of the 2 opponent groups
  Score_diffs[i] <- mean(Tm[E]) - mean(Tm[-E])
}

head(Score_diffs)
summary(Score_diffs)
mean(Score_diffs) # 0.001794507 close to zero
hist(Score_diffs, main="Mean Score difference between games against Eastern vs games against Western Opponents",
     col="light gray", xlab="Mean difference", xlim=c(-55, 55))
#Now display the observed value on the histogram
abline(v = Observed, col = "purple")
#What is the probability (the P value) that a difference this large
#could have arisen with a random subset?
pvalue <- (sum(Score_diffs >= Observed)+1)/(N+1); pvalue # 1
```

This goes to show that the data observed has a significant likelihood to have 
come about by chance. Therefore, there is insufficient evidence to to reject the null hypothesis.




