---
title: "Math E-23C Term Project"
author: "Maria Cristina An, Daniel Lebedinsky, Julio Solis Arce"
date: '2022-05-06'
output: html_document
---

## Glossary
 Rk -- Rank
 G -- Season Game
 W_E -- opponent is from East or West
 Opp -- Opponent
 W_L -- Lakers team won or lost
 Tm -- Lakers team Points
 Opp_pts -- Opponent Points

# Lakers 
 FG_LA -- Field Goals
 FGA_LA -- Field Goal Attempts
 FGpc_LA -- Field Goal Percentage
 X3P_LA -- 3-Point Field Goals
 X3PA_LA -- 3-Point Field Goal Attempts
 X3Ppc_LA -- 3-Point Field Goal Percentage
 FT_LA -- Free Throws
 FTA_LA -- Free Throw Attempts
 FTpc_LA -- Free Throw Percentage
 ORB_LA -- Offensive Rebounds
 TRB_LA -- Total Rebounds
 AST_LA -- Assists
 STL_LA -- Steals
 BLK_LA -- Blocks
 TOV_LA -- Turnovers
 PF_LA -- Personal Fouls

# Opponent
 FG -- Opponent Field Goals
 FGA -- Opponent Field Goal Attempts
 FGpc -- Opponent Field Goal Percentage
 X3P -- Opponent 3-Point Field Goals
 X3PA -- Opponent 3-Point Field Goal Attempts
 X3Ppc -- Opponent 3-Point Field Goal Percentage
 FT -- Opponent Free Throws
 FTA -- Opponent Free Throw Attempts
 FTpc -- Opponent Free Throw Percentage
 ORB -- Opponent Offensive Rebounds
 TRB -- Opponent Total Rebounds
 AST -- Opponent Assists
 STL -- Opponent Steals
 BLK -- Opponent Blocks
 TOV -- Opponent Turnovers
 PF -- Opponent Personal Fouls

## 1. Importing and cleaning data
Reference: https://www.basketball-reference.com/teams/LAL/2020/gamelog/
```{r}
Lakers <- read.csv("Lakers\ 2019-20\ Game\ log.csv"); head(Lakers)
# Convert to proper date format
Dates <- as.Date(Date)
attach(Lakers)
```

## 2. and 3. Summary

```{r}
summary(Lakers)
```

# 4. Number of columns and rows
```{r}
length(Lakers)
nrow(Lakers)
# Creating categorical columns
WonLost <- W_L == "W"; head(WonLost) 
sum(WonLost) # total wins
```

## Graphical Displays

# 1. Barplot of wins vs losses for our team
 This plot shows the number of losses compared to the number wins by our team. 
 The y-axis gives the count of wins and losses. The x-label answers the question
 "did the team win?", FALSE means losing and TRUE means winning.
```{r}
barplot(table(WonLost), main = "Bar Plot of Lakers' 2019 Wins vs Losses", col = "orange", ylab="Count")
```
 

# Barplot of 3 point shots from our team

This plot shows the number of times the team scores a 3-point shot in a game.
How often a 3-pointer occurs ranges from the minimum of 2 to a maximum of 19
in a single game.
```{r}
barplot(table(X3P_LA), main = "Bar Plot of Lakers' Number of 3-Pointers per Game", xlab="Number of 3-pointers", ylab="Frequency", col = "yellow")
```


# Barplot of Opponent blocks 

This plot shows how often an opponent successfully blocks our team in a game.
Most often (the median), an opponent is able to block 3 times and, of course, 
depending on the opponent's abilities, blocks can range from 0 to 9 in a game.
```{r}
barplot(table(BLK), main = "Bar Plot of Lakers' Number of Opponent Blocks per Game", xlab="Number of Blocks", ylab="Frequency", col = "green")
```


## 2. Histogram of 3 point shots

This is a histogram version of the barplot which shows the number of times the team scores a 3-point shot in a game.
The data is the same, and so is the result, a 3-pointer event ranges from the minimum of 2 to a maximum of 19.
```{r}
hist(X3P_LA, xlab = "3 point shots", main = "Histogram of Lakers' 3 Pointers", col = "cyan", breaks=(0:25))
```


# Histogram of Opponent blocks
 
This is a histogram version that shows how often an opponent successfully blocks our team in a game.
Same results, an opponent is able to block 3 times and ranges from 0 to 9 in a game.
```{r}
hist(BLK, main = "Histogram of Lakers' Opponent Blocks", xlab="Number of Blocks", col = "magenta", breaks=(0:12))
```


# Histogram of our team's number of Free Throws

This histogram shows the probability of the number of free throws done by our team in a game.
The range is from a minimum of 7 to a maximum of 33 with a maximum density at 14 and 17.
```{r}
hist(FT_LA, probability=TRUE, main="Histogram of Lakers' Number of Free Throws per Game", xlab="Number of Free Throws", col = "pink", breaks=(0:35)) # free throws from the team per game
```


## 3. Probability density graph overlay on the our team's total number of rebounds in a game

Here we have our team's probability of the total number of rebounds in a game.
Ranging from a minimum of 25 to a maximum of 62, the graph seems like it came from a 
normal distribution. Overlaying a normal distribution shows like it does.
```{r}
mu<-mean(TRB_LA); stdv<- sd(TRB_LA)
hist(TRB_LA, probability=TRUE, main = "Histogram of Lakers' Total Number of Rebounds per Game", ylim=c(0,0.08), xlab="Number of Rebounds", col = "pink", breaks=(20:65)) 
# total rebounds per game  
curve(dnorm(x, mu, stdv), col="purple", add=TRUE, lwd=2) 
#Appears to be a normal distribution. We will compare it with Deciles and a Q-Q plot
Normal10<- qnorm(seq(0.01, 0.99, by=0.1), mean=mu, sd=stdv);
Data10<- quantile(TRB_LA, seq(0.01, 0.99, by=0.1), type=2);
plot(Normal10, Data10)
f<-function(x) x
curve(f, col="red", add=TRUE)
#Our data does not differ much from the Normal Distribution, except in the first decile.
```

# Probability density graph on the our team's scores in a game

We go back to our histogram of our team's score and its probability in a game.
It also seems as if it fits a normal distribution like it shows with a normal distribution overlay.
We can provide more proof later with a chi-square goodness of fit test.
```{r}
mu<-mean(Tm); stdv<-sd(Tm)
hist(Tm,breaks=seq(from=80, to = 150, by =5), probability=TRUE, main = "Histogram of Lakers' Scores per Game", xlab="Scores", col = "light blue") # team's score frequencies
curve(dnorm(x, mu, stdv), col="dark blue", add=TRUE, lwd=2)

#Binning and Chisq test to compare this histogram to a Normal Distribution
dec <- qnorm(seq(0.0, 1, by = 0.1), mu, stdv); dec   #10 bins
Exp <- rep(length(Tm)/10,10); Exp     #expected scores per bin
#Now we can count how many scores are in each bin
binscores <- numeric(10)
for (i in 1:10)  
  binscores[i] <- sum((Tm >= dec[i]) & (Tm <= dec[i+1]) ); binscores
#Test for uniformity using chi square.
Chi2 <- sum((binscores - Exp)^2/Exp); Chi2
#There were 10 bins. We estimated 2 parameters (mu and stdv), which costs two degrees of freedom
#Also we "made the totals match", costing another 1. So there are 10-2-1=7 df.
curve(dchisq(x, df = 7), from = 0, to = 120 )
abline(v=Chi2, col = "red")
#The probability of this chi-square value is relatively large
#The normal distribution was a good model
```

## 4. Contingency tables

This table answers the question "how many times the opponents scored greater than 100 points in a game with our team".
This does not say that the opponent won or not. But in the times that they did not, it only shows that our team scored more and that such a high scoring game would have been a truly awesome event to watch.
```{r}
Opp100 <- Opp_pts > 100; Opp100
table(Opp100)
```

This table shows the opposite of the table above, that despite being the all-around champion during that year, our team scored less than 100 in some games and this table shows that.
```{r}
Less100 <- Tm < 100; Less100
table(Less100)
```

## 1. Permutation test
Our permutation test will be comparing our team's scores in games played vs the opponents from the East and the opponents from the West

H_o: The mean score of the opponents coming from the Eastern Conference is equal
to the mean score of the opponents coming from the Western Conference
H_a: The mean score of the opponents coming from the Eastern Conference is not equal
to the mean score of the opponents coming from the Western Conference

#Data for Eastern and Western Conference teams
How many opponents were from each, and what were their mean scores? 
```{r}
Eastern <- W_E == "E"; head(Eastern) 
East <- sum(Eastern); East # count of Eastern opponents = 31
ScorevsEast <- mean(Tm*Eastern); ScorevsEast # mean team score vs East = 37.03261

Western <- W_E == "W"; head(Western) 
West <- sum(Western); West # Western opponents = 61
ScorevsWest <- mean(Tm*Western); ScorevsWest # mean team score vs West = 76.26087
```
It shows that there are more opponents coming from the West and that their mean score
obviously would have a higher range.

# Total number of games played and differences of mean scores 
This will be our Observed value.
```{r}
EW <- sum(Eastern) + sum(Western); EW # total number of games = 92
Score_diff <- ScorevsEast - ScorevsWest; Score_diff # -39.22826
Observed <- Score_diff; Observed
# Let's see if this score difference is significant
# We repeat 10^6 times
N <- 10^6
Score_diffs <- numeric(N)
for (i in 1:N){
  # Permute West indices
  E <- sample(EW, East, replace = FALSE)
  
  # Get the difference of the 2 opponent groups
  Score_diffs[i] <- mean(Tm[E]) - mean(Tm[-E])
}

head(Score_diffs)
summary(Score_diffs)
mean(Score_diffs) # 0.001794507 close to zero
hist(Score_diffs, main="Mean Score difference between games against Eastern vs games against Western Opponents",
     col="light gray", xlab="Mean difference", xlim=c(-55, 55))
#Now display the observed value on the histogram
abline(v = Observed, col = "purple")
#What is the probability (the P value) that a difference this large
#could have arisen with a random subset?
pvalue <- (sum(Score_diffs >= Observed)+1)/(N+1); pvalue # 1
```

This goes to show that the data observed has a significant likelihood to have 
come about by chance. Therefore, there is insufficient evidence to to reject the null hypothesis.

## The End



